\begin{acknowledgements}

  First and foremost I would like to express my deep gratitude to my supervisor, Henry Wilton, whose guidance, encouragement, positivity and patience have made these four years a rewarding and (mostly) enjoyable experience.
  I could not have hoped for more dedicated supervisor.
  I also thank the many other people in the geometric group theory community, in Cambridge and more broadly, who have taken the time to discuss my work with me, including but certainly not limited to Martin Bridson, Brian Bowditch, Daniel Groves, Panos Papasoglu, Alan Logan, John Mackay and Alex Margolis.
  I am grateful to the organisers of Geometric Group Theory in the South East and of the Young Geometric Group Theory conference series: these events have broadened my understanding of geometric group theory and made me a better mathematician.

  I thank the EPSRC for funding this PhD.
  I also thank Trinity College for being my home in Cambridge for seven years and Churchill College for welcoming me into its community.

  I thank my friends from my time as an undergraduate and graduate student at Cambridge, including James Munro, Fiona Hamey, David Mestel, Jack Smith, Jo Evans, Tom Berrett, Lawrence Barrott, James Kilbane, Nigel Burke and Nicolas Dupr\'e, for their conversation both mathematical and (especially) non-mathematical.
  Living with James Munro and Fiona Hamey has made writing this thesis a more pleasant experience, and I am particularly grateful to them.

  I am also grateful to my friends outside the mathematical community, whose friendship has enriched my life and provided often-needed distraction from mathematics.
  I thank Cambridge Labour Party, and in particular George Owers, Dan Ratcliffe, Dan Swain and Ashley Walsh.

  Finally I thank my family, and in particular my parents, to whom this thesis is dedicated.
  It is thanks to their early encouragement and support that I have reached this point.

\end{acknowledgements}

